---
title: "Statistical Inference Part 1 - Simulation"
output:
  pdf_document:
    fig_height: 4
    toc_depth: 1
  html_document:
    keep_md: yes
date: "Sunday, July 26, 2015"
---
## Overview
In this course project we will investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda (1/$\lambda$) and the standard deviation is also 1/lambda (1/$\lambda$). We will set lambda = 0.2 for all of the simulations. We will investigate the distribution of averages of 40 exponentials.

##Simulations
Let's do a thousand simulated averages of 40 exponentials and then 10000 simulated averages of 40 exponentials.
```{r}
set.seed(3) # This is to ensure the same random sample is selected each time
lambda <- 0.2 #set our lamda
number_of_simulations <- 1000
sample_size <- 40
simulation <- matrix(rexp(number_of_simulations*sample_size, rate=lambda), 
                     number_of_simulations, sample_size)
row_means <- rowMeans(simulation)
```

##The distribution of sample means
```{r echo=FALSE}
par(mar=c(2,2,3,2))
# plot the histogram of averages
hist(row_means, breaks=50, prob=TRUE,
     main="Distribution of means of samples (1000),
     drawn from exponential distribution with lambda=0.2",
     xlab="")
# density of the averages of samples
lines(density(row_means), col="blue", lwd=2)
# theoretical center of distribution
abline(v=1/lambda, col="red", lwd=2, lty=2)
abline(v=mean(row_means), col="blue", lwd=2)
# theoretical density of the averages of samples
xfit <- seq(min(row_means), max(row_means), length=100)
yfit <- dnorm(xfit, mean=1/lambda, sd=(1/lambda/sqrt(sample_size)))
lines(xfit, yfit, pch=22, col="red", lty=2, lwd =2)
# add a legend to the top right
legend('topright', c("simulation", "theoretical"), lty=c(1,2), col=c("blue", "red"))
```
As you can see from the graph below, the calculated distribution of means of random sampled exponantial distributions, overlaps quite nice with the normal distribution with the expected values based on the given lamba
The graph shows the the sample (simulated in blue) and theoretical (in red) distribution. As you can see the graph is approxiametly normally distributed. The means are very close together (centre of distribution) as demonstrated by the red and blue vertical lines for theoretical and simulated lines respectively. 

At 10000 simulations the means are almost identical while the distributions are also almost the same.

```{r echo=FALSE}
set.seed(453) # This is to ensure the same random sample is selected each time
lambda <- 0.2 #set our lamda
number_of_simulations <- 10000
sample_size <- 40
simulation <- matrix(rexp(number_of_simulations*sample_size, rate=lambda), 
                     number_of_simulations, sample_size)
row_means <- rowMeans(simulation)

# plot the histogram of averages
hist(row_means, breaks=50, prob=TRUE,
     main="Distribution of means of samples(10000),
     drawn from exponential distribution with lambda=0.2",
     xlab="")
# density of the averages of samples
lines(density(row_means), col="blue", lwd=2)
# theoretical center of distribution
abline(v=1/lambda, col="red", lwd=2, lty=2)
abline(v=mean(row_means), col="blue", lwd=2)
# theoretical density of the averages of samples
xfit <- seq(min(row_means), max(row_means), length=100)
yfit <- dnorm(xfit, mean=1/lambda, sd=(1/lambda/sqrt(sample_size)))
lines(xfit, yfit, pch=22, col="red", lty=2, lwd =2)
# add a legend to the top right
legend('topright', c("simulation", "theoretical"), lty=c(1,2), col=c("blue", "red"))
```

## Sample Mean versus Theoretical Mean

The expected mean $\mu$ of a exponential distribution of rate $\lambda$ is 

$\mu= \frac{1}{\lambda}$ 

```{r}
mu <- 1/lambda
mu
```

Let $\bar X$ be the average sample mean of 1000 simulations of 40 randomly sampled exponential distributions.

```{r}
meanOfMeans <- mean(row_means)
meanOfMeans
```

As you can see the expected mean and the avarage sample mean are very close. This is also shown in the graph above with the expected mean in Red vertical line and the sample mean in the Blue verical line. 


## Sample Variance versus Theoretical Variance

The expected standard deviation $\sigma$ of a exponential distribution of rate $\lambda$ is 

$\sigma = \frac{1/\lambda}{\sqrt{n}}$ 


```{r}
sd <- 1/lambda/sqrt(sample_size)
sd
```

The variance $Var$ of standard deviation $\sigma$ is

$Var = \sigma^2$ 

```{r}
Var <- sd^2
Var
```

Let $Var_x$ be the variance of the average sample mean of 1000 simulations of 40 randomly sampled exponential distribution, and $\sigma_x$ the corresponding standard deviation.
```{r}
sd_x <- sd(row_means)
sd_x

Var_x <- var(row_means)
Var_x
```

As you can see the standard deviations are very close
Since variance is the square of the standard deviations, minor differnces will we enhanced, but are still pretty close.
